PhD Proposal

### 2022-05-02 abstract/abstract.tex
continuo abstract

\documentclass[a4paper,11pt, oneside]{article}
\usepackage[left=15mm, right=15mm, top=20mm, bottom=20mm]{geometry}
\usepackage[utf8]{inputenc}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{graphics}
\usepackage{amsfonts}
\usepackage{listings}
\usepackage{amsmath}
\usepackage{amstext}
\usepackage{engrec}
\usepackage{rotating}
\usepackage{verbatim}
\usepackage[safe,extra]{tipa}
% \usepackage{showkeys}
\usepackage{multirow}
\usepackage{hyperref}
\usepackage{microtype}
\usepackage{fontspec}
\usepackage{enumerate}
\usepackage{physics}
\usepackage{braket}
\usepackage{mhchem}
\usepackage{marginnote}
\usepackage{pgfplots}
\usepackage{cancel}
\usepackage{polynom}
\usepackage{booktabs}
\usepackage{enumitem}
\usepackage{framed}
\usepackage{pdfpages}
\usepackage{pgfplots}
\usepackage{algorithm}
% \usepackage{algpseudocode}
\usepackage[cache=false]{minted}
\usepackage{mathtools}
\usepackage[noend]{algpseudocode}

@@

\usepackage{tikz}\usetikzlibrary{er}\tikzset{multi  attribute /.style={attribute
    ,double  distance =1.5pt}}\tikzset{derived  attribute /.style={attribute
    ,dashed}}\tikzset{total /.style={double  distance =1.5pt}}\tikzset{every
  entity /.style={draw=orange , fill=orange!20}}\tikzset{every  attribute
  /.style={draw=MediumPurple1, fill=MediumPurple1!20}}\tikzset{every
  relationship /.style={draw=Chartreuse2,
    fill=Chartreuse2!20}}\newcommand{\key}[1]{\underline{#1}}
\usetikzlibrary{arrows.meta}
\usetikzlibrary{decorations.markings}
\usetikzlibrary{arrows,shapes,backgrounds,petri}
\tikzset{
  place/.style={
    circle,
    thick,
    draw=black,
    minimum size=6mm,
  },
  transition/.style={
    rectangle,
    thick,
    fill=black,
    minimum width=8mm,
    inner ysep=2pt
  },
  transitionv/.style={
    rectangle,
    thick,
    fill=black,
    minimum height=8mm,
    inner xsep=2pt
  }
} 
\usetikzlibrary{automata,positioning,chains,fit,shapes}
\usetikzlibrary{circuits.logic.US}
\usetikzlibrary{positioning}
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhead[LE,RO]{\slshape \rightmark}
\fancyhead[LO,RE]{\slshape \leftmark}
\fancyfoot[C]{\thepage}
\usepackage[usenames,dvipsnames]{pstricks}
\usepackage{epsfig}
\usepackage{pst-grad} % For gradients
\usepackage{pst-plot} % For axes
\usepackage[space]{grffile} % For spaces in paths
\usepackage{etoolbox} % For spaces in paths
\makeatletter % For spaces in paths
\patchcmd\Gread@eps{\@inputcheck#1 }{\@inputcheck"#1"\relax}{}{}
\makeatother
\usepackage{lipsum}
\DeclareSymbolFont{symbolsC}{U}{txsyc}{m}{n}
\DeclareMathSymbol{\strictif}{\mathrel}{symbolsC}{74}
\title{PhD Proposal}
\author{Davide Cozzi, 829827,
  \href{mailto:d.cozzi@campus.unimib.it}{d.cozzi@campus.unimib.it}} 
\date{}

@@

\section*{Introduction}

@@

the field of algorithmics and bioinformatics. For example the interest in such
problems is due to the need to align sequences or search for specific patterns
within the \textit{DNA}. \\
In this context, a large number of data structure algorithms have been

@@

  transform (\textit{BWT})} thanks to the studies of Ferragina and Manzini who
proposed its use together with the so-called FM-index \cite{fmindex}.\\
In recent years, in the field of bioinformatics, there is a change of
interest. While until a few years ago the research was focused on the study of
the \textit{genome}, we are now starting to deepen the topic of
\textit{pangenome}. Currently, the need to take into

@@

In this context, various algorithms and various data structures have been
implemented in order to study the so called \textit{haplotypes} and
\textit{genotypes}. Briefly we could define \textit{haplotypes} as a combination
of allelic variants, inherited from a parent. Instead we can define the
\textit{genotype} as the complete set of genes contained in the DNA. From a
biological point of view it is indeed interesting to note that each individual
shares about 99\% of the genetic code, while the remaining 1\% differs mostly
with \textbf{Single Nucleotide Polymorphisms (\textit{SNPs})}, i.e. variations
of a single nucleotide in a precise \textit{locus} of the DNA\\ 
Thanks to the last developments in sequencing technologies, which have led both

@@

\textit{pangenome graph}, replacing the old representation of the single genome
by a single sequence. \\
One of the most important data structure developed in order to handle the study
of haplotypes sequences is the \textbf{positional Burrows-Wheeler transform
  (\textit{PBWT})}, proposed by Durbin in 2014 \cite{pbwt_durbin}. With this
particular data structure it was possible both to study possible matches inside
a haplotype panel and to search for set maximal matches between an external
haplotype and a panel, as well as the set maximal matches inside the panel
itself. Furthermore, for example, variants have also been studied for the 
management of the multiallelic case \cite{mpbwt}. \\
In 2021 Rossi et al. proposed \textit{MONI} as a data structure to handle a
\textbf{run-length version of BWT (\textit{RLBWT})} with the ultimate intention
of indexing and using multiple genomes as a reference \cite{moni}. Together with
this data structure the authors proposed the concept of \textit{matching
  statistics} in order to efficiently compute the matches between a pattern and
a text. A recent improvement, regarding the \textit{RLBWT}, has been made
through the implementation of \textit{PHONI} \cite{phoni}, where where the
so-called \textit{longest-common-extension (LCE) queries} are used. \\
During the development of my master's thesis I worked in collaboration with the
authors of MONI in order to create a run length encoded variant of the PBWT. In
this work I used practically all concepts applied to the ``classic''
\textit{BWT} in \textit{MONI} and \textit{PHONI}. \\
During my PhD I will focus my studies on the development of new algorithms in
various topics such as variants of the \textit{PBWT} (and any related uses),
\textit{haplotyping/genotyping} or solutions concerning the study of
\textit{SNPs}.It is also my intention to deepen the more experimental themes
relating to pattern matching, in detail the new developments on BWT and new
indexing structures, as well as the theme of \textit{succinct data structures},
with particular attention to the use of \textit{bitvectors}. 
\section*{State of the art}
I will now present a brief overview of the main algorithms, data structures,
methods etc that will be the core of my studies during my PhD. 
\subsection*{BWT}
The \textbf{Burrows-Wheeler Transform (\textit{BWT})} \cite{bwt} was introduced
in 1994 in order to compress texts but it has had then wide use in
bioinformatics, above all thanks to the already cited \textit{FM-index}. Given a
text $T$, \$-terminated, such that $|T|=n$, we can define the $BWT_T$, denoting
with $SA_T$ the \textit{suffix array} of $T$, as:
$BWT_T[i] = T[SA_T[i]-1]$, if $SA_T[i]>0$, and $BWT_T[i] = \$ $ otherwise. Less
formally we can say that $BWT_T[i]$ is the character that precedes the $i$-th
suffix in the lexicographically order. It is important to note that this
transform is reversible so we can reconstruct the text $T$ from its transform
$BWT_T$ using the so-called \textbf{LF-mapping}. Given $BWT_T$ and an array,
called $F_T$, with all the characters of $T$ in the lexicographically order, we
can say that, thanks to the \textit{LF-mapping}, the $j$-th occurrence of a
certain character in $BWT_T$ corresponds to the $j$-th occurrence of the same
character in $F_T$, so we can reconstruct $T$ starting from its last character
\$. With the use of the \textit{LF-mapping} we can perform the
\textit{backward-search} in order to use the $BWT_T$ to look for a pattern $P$
within $T$. This can be done efficiently thank to the \textit{FM-index} which
consists of two functions. The first one is $C$ function, such that, given an
alphabet $\Sigma$ (that includes the ending character \$),
$C:\Sigma\to[1,n]$. This function, given a character $\sigma\in \Sigma$ returns
the number of occurrences of characters lexicographically than the one given as
argument in $T$. The second one is the $Occ$ function,
$Occ:\Sigma\times[1,n]\to[1,n]$, has as arguments a character $\sigma\in\Sigma$
and an index $i$ of $BWT_T$ and returns the count of occurrences of $\sigma$ in
$BWT_T[1,i]$ (\textbf{CONTROLLARE E RISCRIVERE MEGLIO}).\\
The use of \textit{BWT} has allowed the construction of efficient algorithms
both in the field of pattern matching and in that of sequence alignment. 
\subsection*{Bitvectors}
\textit{Bitvectors} are ones of the most important data structure when
mentioning succinct data structures. \\

@@

operations possible on Boolean arrays, such as \textit{random access} in
\textit{constant time} (\textbf{DA VERIFICARE}). More in detail the rank
function allows you to calculate how many values of 1 are up to a certain
index. Instead the select function allows to obtain the index of any one present
in the bitvector. Formally, given a bitvector $B$, such that $|B|=n$, and given
an index $i$, such that $0\leq i<n$, we can define $rank_B(i)=\sum_{k=0}^{k<i}
B[k]$. Instead, about the select function, given an integer $i$, such that
$0<i\leq rank_B(n)$, where $n=|B|$,  we can define $select(i)=\min\{j \,| \,\,
rank_B(j+1)=i\}$.\\
From a purely theoretical point of view, with the additional cost of
$\mathcal{O}(n)$ bits in memory, these two operations can be supported with
constant time.  In more practical terms there are several implementations of the
same within \textbf{SDSL (\textit{Succint Data Structures Library})}
\cite{sdsl}, one of the most important C++ library used in bioinformatics. As

@@

operations varies (usually only one of the two is in constant time) and the
amount of additional bits needed.\\
An example of the use of bitvectors is to track the runs in the run-length
encoded implementations of \textit{BWT} and \textit{PBWT}, where we put one at
each head of run.

\subsection*{RLBWT}
Speaking of pangenome, linear indexing via FM-index is no longer the best
solution as it does not handle the large repetitions present in this new type of
sequences in the best possible way. In 2005 M\"{a}niken and Navarro defined

@@

$T$, $RLBWT_T$ is a rappresentation of $BWT_T$ with a compact rappresentation of
consecutive equal characters, the so-called \textit{runs}. With this change of
perspective the algorithms have gone from being linear over the length of the
text, $n$, to being linear over the number of runs, $r$. \\
The new indexing method, which was then introduced by Gagie et
al. \cite{rindex}, was called \textbf{r-index} and corresponds to the
\textit{RLBWT} and a SA sampling at the begin and at the end of every run. The
algorithm for querying through the \textit{RLBWT} take advantage of other
methods, such as the use of \textbf{thresholds} (minimum
\textit{LCP} value between two consecutive runs of the same character) in
\textit{MONI}, and the use of \textbf{longest common extension (\textit{LCE})
  query} (to compute the right equal common extension between two position in
the text) cd\cite{lce} and \textbf{straight-line programs (\textit{SLP})}
\cite{slp} (for \textit{random access} and \textit{lce queries} over the
grammar-compressed text) in \textit{PHONI}. Both these solutions are used for
the computation of the \textbf{matching statistics (\textit{MS})}. Given a
pattern $P$ and a text $T$ the \textit{MS} of $P$ in respect to $P$ is an array
$M$ of pairs position/length, $|M|=|P|$, such that
$T[M[i].pos:M[i].pos+M[i].len-1=P[i:i+M[i].len]-1]$ and $P[i:i+M[i].len]$ does
not occur in $T$. Given \textit{MS} we can compute every \textbf{Maximal Exact
  Match (\textit{MEM})} of a pattern in a text. Given a text $T$ and a pattern
$P$ a substring of the pattern $P[i : i+l-1]$, of length $l$, is a \textit{MEM}
of $P$ in $T$ if $P[i:i+l-1]$ is a substring of $T$ but but neither
$P[i-1:i+l-1]$ nor $P[i:i+l]$ are. At the end, using a particular function
called $\varphi$ (and $\varphi^{-1}$), based on the use of the \textbf{inverse
  suffix array (\textit{ISA})} as well, it was possible to start from the
starting positon of a match extracted 
by \textit{MS} and find all the starting positions of all the copies of $P$ in
$T$. Infact, formally, given a starting position $p$, we have
$\varphi(p)=SA[ISA[p]-1]$, $NULL$ if $ISA[p]=0$, and
$\varphi^{-1}(p)=SA[ISA[p]+1]$, $NULL$ if $ISA[p]=|T|-1$, where $ISA[i]=j$ iff
$SA[j]=i$.\\   
Thanks to these and other methods it was possible to perform pattern matching

@@

a pangenomic context. 
\subsection*{PBWT}
Based on the theories of BWT Durbin, in 2014, he devised the positional
\textbf{Burrows–Wheeler transform (\textit{PBWT}}) \cite{pbwt}, in order to
solve the problem of pattern matching on panels (matrices) of haplotypes,
denoted by $X$. In
detail we are talking about a panel with $M$ haplotypes and $N$ biallelic
sites. This data structure is based on a reversed-prefix ordering at each column
$k$ that produces two different multidimensional arrays. The first one is called
\textbf{prefix array}, denoted by $a$, and, for each column $k$, contains, for
each position $i$, the haplotype of index $m$ in the original panel. More
formally we can say that $a_k[i]=m$ iff $X_m$ is the $i$-th haplotype in the
reversed-prefix ordering at column $k$. Note that $x_m$ such that $a_k[i]=m$
could be denoted by $y_i^k$. The second bidimensional array is
called \textbf{divergence array}, denoted by $d$, and indicates the index of the
starting column of the longest common suffix, ending in column $k$, between a
row and its previous one, at reversed-prefix ordering at column $k$. More
formally we can define $d_k[i]=h$ iff $h$ is the smallest column index such that
$y_i^k[h,k)=y_{i-1}^k[h,k)$. (\textbf{RISCRIVERE MEGLIO})\\
Thanks to these two bidimensional arrays it is possible to compute all matches

@@

in linear time, all set-maximal matches (which we could also call ``MEMs'') from
a new sequence $z$ to $X$ etc$\ldots$ 
\section*{Research goals}
For my master's thesis  I therefore tried to combine the ideas related to the
\textit{RLBWT} with those of the \textit{PBWT}, creating the \textbf{RLPBWT}. In
order to obtain this result I have to rethink the concept of \textit{Matching
  Statistics} for \textit{PBWT}, how to compute the \textit{SLP} for the panel,
how to use \textit{thresholds}/\textit{LCE queries}, how to obtain the same
behaviour of the $\varphi$ function etc$\ldots$\\
In detail, regarding \textit{MS}, instead of the \textit{pos} we track a
\textit{row} of the panel, regarding thresholds we use the index inside a run in
a column where we have the minimum value of divergence array and, regarding the
\textit{SLP}, we ``stretch'' the reverse panel (from the right to the left) in
in order to make \textit{LCE queries} possible. Instead, regarding the $\varphi$
function, I have implemented a new simple data structure to identify which row
is above and which row is below each row in the panel permutated via the
prefix array and stored in a run-length compressed way. (\textbf{RISCRIVERE
  MEGLIO ED EVENTUALMENTE APPROFONDIRE}).\\
Obviously there are some limitations, such as the study of biallelic panels
only and the lack of management of any missing data, which are very frequent in
the case of real data. First of all it will be interesting to implement a new

>>>>

\documentclass[a4paper,11pt, oneside]{article}
\usepackage[left=15mm, right=15mm, top=15mm, bottom=20mm]{geometry}
\usepackage[utf8]{inputenc}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{graphics}
\usepackage{amsfonts}
\usepackage{listings}
\usepackage{amsmath}
\usepackage{amstext}
\usepackage{engrec}
\usepackage{rotating}
\usepackage{verbatim}
\usepackage[safe,extra]{tipa}
% \usepackage{showkeys}
\usepackage{multirow}
\usepackage{hyperref}
\usepackage{microtype}
\usepackage{fontspec}
\usepackage{enumerate}
\usepackage{physics}
\usepackage{braket}
\usepackage{mhchem}
\usepackage{marginnote}
\usepackage{pgfplots}
\usepackage{cancel}
\usepackage{polynom}
\usepackage{booktabs}
\usepackage{enumitem}
\usepackage{framed}
\usepackage{pdfpages}
\usepackage{pgfplots}
\usepackage{algorithm}
% \usepackage{algpseudocode}
\usepackage[cache=false]{minted}
\usepackage{mathtools}
\usepackage[noend]{algpseudocode}

@@

\usepackage{tikz}\usetikzlibrary{er}\tikzset{multi  attribute /.style={attribute
    ,double  distance =1.5pt}}\tikzset{derived  attribute /.style={attribute
    ,dashed}}\tikzset{total /.style={double  distance =1.5pt}}\tikzset{every
  entity /.style={draw=orange , fill=orange!20}}\tikzset{every  attribute
  /.style={draw=MediumPurple1, fill=MediumPurple1!20}}\tikzset{every
  relationship /.style={draw=Chartreuse2,
    fill=Chartreuse2!20}}\newcommand{\key}[1]{\underline{#1}}
\usetikzlibrary{arrows.meta}
\usetikzlibrary{decorations.markings}
\usetikzlibrary{arrows,shapes,backgrounds,petri}
\tikzset{
  place/.style={
    circle,
    thick,
    draw=black,
    minimum size=6mm,
  },
  transition/.style={
    rectangle,
    thick,
    fill=black,
    minimum width=8mm,
    inner ysep=2pt
  },
  transitionv/.style={
    rectangle,
    thick,
    fill=black,
    minimum height=8mm,
    inner xsep=2pt
  }
} 
\usetikzlibrary{automata,positioning,chains,fit,shapes}
\usetikzlibrary{circuits.logic.US}
\usetikzlibrary{positioning}
\usepackage{fancyhdr}
% \pagestyle{fancy}
% \fancyhead[LE,RO]{\slshape \rightmark}
% \fancyhead[LO,RE]{\slshape \leftmark}
\fancyfoot[C]{\thepage}
\usepackage[usenames,dvipsnames]{pstricks}
\usepackage{epsfig}
\usepackage{pst-grad} % For gradients
\usepackage{pst-plot} % For axes
\usepackage[space]{grffile} % For spaces in paths
\usepackage{etoolbox} % For spaces in paths
\makeatletter % For spaces in paths
\patchcmd\Gread@eps{\@inputcheck#1 }{\@inputcheck"#1"\relax}{}{}
\makeatother
\usepackage{lipsum}
\DeclareSymbolFont{symbolsC}{U}{txsyc}{m}{n}
\DeclareMathSymbol{\strictif}{\mathrel}{symbolsC}{74}
\title{PhD Proposal}
\author{Davide Cozzi, 829827,
  \href{mailto:d.cozzi@campus.unimib.it}{d.cozzi@campus.unimib.it}} 
\date{}
\makeatletter
\renewcommand{\paragraph}{%
  \@startsection{paragraph}{4}%
  {\z@}{1.25ex \@plus 1ex \@minus .2ex}{-1em}%
  {\normalfont\normalsize\bfseries}%
}
\makeatother

@@

\subsubsection*{Introduction}

@@

the field of algorithmics and bioinformatics. The interest in such
problems is due to the need to align sequences or to search for specific
patterns within the \textit{DNA}. 
In this context, a large number of data structures and algorithms have been

@@

  transform (\textit{BWT})}, thanks to the studies of Ferragina and Manzini, who
proposed its use together with indexing via \textit{FM-index}.\\
Furthermore, in recent years, there has been a change of interest in the field
of bioinformatics. Until a few years ago the research was focused on the
study of a \textbf{linear sequence of a genome} while now the 
researchers are beginning to deepen the topic of the \textbf{pangenome}, which
term was 
introduced by Tettelin in 2005. In fact, the need to take into 

@@

Thanks to the last developments in sequencing technologies, which had led both

@@

\textbf{pangenome graph}. Furthermore, the new amount of sequences has led to
new algorithms regarding the problem of \textit{pattern matching}. In 2021,
Rossi et al. proposed \textit{MONI} as a data structure to handle a 
\textbf{run-length encoded version of BWT (\textit{RLBWT})}, with the ultimate
intention of indexing and using multiple genomes as a reference
\cite{moni}. Together with this data structure, the authors proposed the concept
of \textbf{matching statistics (\textit{MS})} in order to efficiently compute
the matches between a pattern and a text. A recent improvement has been made
through the implementation of \textit{PHONI} \cite{phoni}, where the
\textbf{longest-common-extension (\textit{LCE}) queries} in order to further
optimize the pattern search.\\ 
From a biological point of view, it is interesting to note that, as it has been
pointed out by the study of \textit{1000 Genome Project}, there are over 88
million variants between those human genomes. Among these variants 84.7 million
are \textbf{Single 
  Nucleotide Polymorphisms (\textit{SNPs})}, 3.6 million are \textbf{short
  insertions/deletions (\textit{indel})} and 60000 are \textbf{structural
  variants}, 
involving more than 
50 nucleotides. All of them are now a limit to the traditional use of a
\textit{linear sequence of a genome}.\\
Against this background, various algorithms and data structures have
been implemented in order to study \textit{haplotypes} and
\textit{genotypes}, such as the \textbf{genotyping variants
  problem}. Briefly, we could define \textit{haplotypes} as a combination 
of allelic variants, each one inherited from a parent. Instead, the
\textit{genotype} is the complete set of genes contained in the DNA. So, from
2005, publication of the \textbf{genome-wide association studies
  (\textit{GWAS})} has begun. The goal of this type of studies is to screen the 
\textit{pangenome} looking for associations between genetic variants, in order,
for example, to study also deseases outocomes.
In this particular historical period it is also impossible not to mention the
viruses. In fact, by nature, during infections, viruses replicate a lot but
often in a not perfect way, thus producing many inexact clones, referred as
\textbf{viral haplotypes} which, taken together, form the \textbf{viral
  pangenome}. The identification of all this haplotypes is crucial both for the
effective study of the spread of the virus and for the production of effective
drugs, in a content of high pharmacological resistance. \\
One of the most important data structure, developed in order to handle the study
of haplotypes sequences, is the \textbf{positional Burrows-Wheeler transform
  (\textit{PBWT})}, proposed by Durbin in 2014 \cite{pbwt_durbin}. Using this
particular data structure (which will be described below), it is possible to
study efficiently a collection of haplotypes but only in the bi-allelic
case. Furthermore variants of the \textit{PBWT} have been studied for
handling the multiallelic case. The use of the \textit{PBWT} is found in many
software for the study of haplotypes and in various genotype imputation methods,
that are studies that infers unobserved genotypes in a sample of individuals
\cite{impute5}.\\
During the development of my master thesis, I worked to create a run length
encoded variant of the \textit{PBWT}, the \textbf{RLPBWT}, using and adapting
the various theories developed for the \textit{RLBWT}, in collaboration with the
authors of \textit{MONI} and \textit{PHONI}. In this context, my PhD
is going to be focused on the development of new algorithms in
various topics related to open problems in the study of the
\textit{haplotyping/genotyping}, of the \textit{genome 
  variants} and of the imputation issues related. My intention is also to deepen
experimental themes  
of \textit{pattern matching} and of the \textit{pangenome graph}, in detail the
new developments on \textit{BWT} and indexing structures, as well as
\textit{succinct data structures}, with particular attention to the use of
\textit{bitvectors}.
\subsubsection*{State of the art}
Now I present a brief overview of the main algorithms, data structures,
methods etc $\ldots$ that will be the core of my studies during PhD. 
% \paragraph*{BWT}
% The \textbf{Burrows-Wheeler Transform (\textit{BWT})} was introduced
% in 1994 in order to compress texts but it has been used widely
% in bioinformatics, above all thanks to the already cited
% \textit{FM-index}. Given a text $T$, \$-terminated, such that $|T|=n$, we can
% define, denoting with $SA_T$ the \textit{suffix array} of $T$, the $BWT_T$ as:
% $BWT_T[i] = T[SA_T[i]-1]$, if $SA_T[i]>0$, and $BWT_T[i] = \$ $ otherwise. Less
% formally, we can say that $BWT_T[i]$ is the character that precedes the $i$-th
% suffix in the lexicographical order. It is important to note that this
% transform is reversible, so we can reconstruct the text $T$ from its transform
% $BWT_T$ using the \textbf{LF-mapping}. Given $BWT_T$ and an array,
% called $F_T$, with all the characters of $T$ in the lexicographical order, we
% can say that, thanks to the \textit{LF-mapping}, the $j$-th occurrence of a
% certain character in $BWT_T$ corresponds to the $j$-th occurrence of the same
% character in $F_T$. Thank to this we can reconstruct $T$ starting from its last
% character 
% \$. With the use of the \textit{LF-mapping}, we can perform the
% \textit{backward-search} in order to use the $BWT_T$ to look for a pattern $P$
% within $T$. This can be done efficiently thank to the \textit{FM-index} which
% consists of two functions. The first one is $C$ function, such that, given an
% alphabet $\Sigma$ (that includes the ending character \$),
% $C:\Sigma\to[1,n]$. This function, given a character $\sigma\in \Sigma$, returns
% the number of occurrences of characters lexicographically minor in $T$ than the
% one given as argument. The second one is the $Occ$ function,
% $Occ:\Sigma\times[1,n]\to[1,n]$, which has a character $\sigma\in\Sigma$ 
% and an index $i$ of $BWT_T$ as arguments and returns the count of occurrences of
% $\sigma$ in $BWT_T[1,i]$.\\
% The use of \textit{BWT} has allowed the construction of efficient algorithms
% both in the field of pattern matching and sequence alignment. 
\paragraph*{Bitvectors}
\textbf{Bitvectors} are one of the most important data structure when
mentioning \textit{succinct data structures}. \\

@@

operations on boolean arrays, such as \textit{random access} in
\textit{constant time}. More in detail, the \textit{rank
function} allows to calculate how many occurrences of one are up to a certain
index. Instead, the \textit{select function} allows to obtain the index of every
one present in the \textit{bitvector}. Formally, given a bitvector $B$, such
that $|B|=n$, and given an index $i$, such that $0\leq i<n$, we can define
$rank_B(i)=\sum_{k=0}^{k<i} B[k]$. Instead, about the select function, given an
integer $i$, such that $0<i\leq rank_B(n)$, where $n=|B|$, we can define
$select(i)=\min\{j \,| \,\, rank_B(j+1)=i\}$.\\ 
From a theoretical point of view these two operations can be supported in
\textit{constant time}, with the additional cost of $\mathcal{O}(n)$ bits in
memory. In more practical terms, there are several implementations of the 
same within \textbf{SDSL (\textit{Succint Data Structures Library})}, one of the
most important C++ library used in bioinformatics. As 

@@

operations varies (usually only one of the two is in constant time) as well as
the amount of additional bits needed. An example of the use of bitvectors is
tracking the runs in the run-length encoded implementations of \textit{BWT} and
\textit{PBWT}, where we put one at each head of run, allowing fast operations
along the runs themselves. 
\paragraph*{RLBWT}
The \textbf{Burrows-Wheeler Transform (\textit{BWT})} was introduced
in 1994 in order to compress texts but it has been used widely
in bioinformatics, above all thanks to the already cited
\textit{FM-index}.\\
Speaking of \textit{pangenome}, linear indexing via FM-index is no longer the
best solution as it does not handle the large repetitions there are in this new
type of sequences. In 2005 M\"{a}niken and Navarro defined 

@@

$T$, $RLBWT_T$ is a rappresentation of $BWT_T$ with a compact storage of
consecutive equal characters, the so-called \textit{runs}. With this new
perspective, the algorithms have changed from being linear over the length of
the text, $n$, to being linear over the number of runs, $r$, so sub-linear over
the length of the text. \\ 
The new indexing method, introduced by Gagie et al., was called \textbf{r-index}
and it corresponds to the \textit{RLBWT} plus the \textit{suffix array sampling}
at the beginning and at the end of every run. The algorithm for querying through
the \textit{RLBWT} takes advantage of other methods, such as the use of
\textbf{thresholds} (minimum \textit{LCP} value between two consecutive runs of
the same character) in \textit{MONI}, and the use of \textbf{longest common
  extension (\textit{LCE}) query} (to compute the right equal common extension
between two position in the text) in \textit{PHONI}. Both solutions use
\textbf{straight-line programs (\textit{SLP})}, for \textit{random 
  access} in \textit{MONI} and for \textit{Longest Common extensions (LCE)
  queries} in \textit{PHONI}. 
The purpose of the two projects is computation of the \textbf{matching
  statistics (\textit{MS})}. Given a pattern $P$ and a text $T$, the \textit{MS}
of $P$ in respect to $T$ is an array $M$ of pairs position/length ($pos$/$len$),
$|M|=|P|$, such that $T[M[i].pos:M[i].pos+M[i].len-1]=P[i:i+M[i].len]-1]$ and
$P[i:i+M[i].len]$ does not occur in $T$. Given \textit{MS}, we can compute every
\textbf{Maximal Exact Match (\textit{MEM})} of a pattern in a text. Given a
text $T$ and a pattern $P$, a substring of the pattern $P[i : i+l-1]$, of length
$l$, is a \textit{MEM} of $P$ in $T$ if $P[i:i+l-1]$ is a substring of $T$ but
neither $P[i-1:i+l-1]$ nor $P[i:i+l]$ are, so if the substring cannot be
extended either right or left. Furthermore, using a particular
function called $\varphi$ (and and its inverse $\varphi^{-1}$), based on the use
of the \textbf{inverse suffix array (\textit{ISA})}, it has been possible to
find all starting positions of all copies of $P$ in $T$ from starting position
of the match extracted by \textit{MS}, quickly calculating, given a position $p$
in $SA$, the previous and next position in the suffix array. \\
% More formally, given a starting position
% $p$, $\varphi(p)=SA[ISA[p]-1]$, $NULL$ if $ISA[p]=0$, and
% $\varphi^{-1}(p)=SA[ISA[p]+1]$, $NULL$ if $ISA[p]=|T|-1$, where $ISA[i]=j$ iff
% $SA[j]=i$.\\
Thanks to these and other methods, it was possible to perform pattern matching

@@

a pangenomic context. In fact, most studies in the field of bioinformatics start
with the resolution of pattern matching problems and, as a direct consequence,
of alignment problems. 
\paragraph*{PBWT}
Based on the theories of the \textit{BWT}, in 2014, Durbin devised the
\textbf{positional Burrows–Wheeler transform (\textit{PBWT})}, in order to
solve the problem of pattern matching on panels (matrices) of haplotypes. In
detail, he analyzed a panel $X$ with $M$ haplotypes and $N$ biallelic
sites. This data structure is based on a \textit{reversed-prefix ordering} at
each column $k$ that produces two different multidimensional arrays. The first
one is the set of the \textbf{prefix arrays}, denoted by $a$, which contains the
index of the haplotype $m$ in the original panel, for each column $k$ and for  
each position $i$ of $a_k$. More formally, we can say that $a_k[i]=m$ iff $X_m$
is the $i$-th haplotype in the reversed-prefix ordering at column $k$. We can
note $x_m$, such that $a_k[i]=m$, could be denoted by $y_i^k$. The second
bidimensional array is the set of the \textbf{divergence arrays}, denoted by
$d$, which indicates the index of the starting column of the longest common
suffix, ending in column $k$, between a row and its previous one, at
reversed-prefix ordering at column $k$. More formally, we can define $d_k[i]=h$
iff $h$ is the smallest column index such that $y_i^k[h,k)=y_{i-1}^k[h,k)$. \\
Thanks to these two bidimensional arrays, it is possible to compute all matches

@@

in linear time and all set-maximal matches (which we could also call ``MEMs'')
from a new sequence $z$ to $X$ in $\mathcal{O}(M^2N)$.
Despite the fact that \textit{PBWT} has been poorly regarded in the scientific
community in the early years of its development, there has been a growth in
research based on it, both in terms of variant design, such as the already
mentioned multiallelic version or the \textbf{dynamic PBWT (\textit{d-PBWT}}),
and in terms of its use to study haplotype panels.\\
A first example could be found in the paper of Alanko et al., published in
2021, where the authors used the \textit{PBWT} to look for \textit{maximal
  perfect haplotype block}, within a haplotypes panel. These types of
algorithms are essential for the identification of genomic regions that show
signatures of natural selection.\\
Another interesting paper is the one by Wlliams and Mumey, published in 2020,
who also studied \textit{maximal perfect haplotype blocks}, but with the
addition of \textit{missing data}, handled with the help of wildcards, using the
\textit{PBWT}. \\ 
A very intriguing tool, talking about \textit{GWAS}, to be cited is
\textbf{IMPUTE5}, proposed by Rubinacci et al. in 2020. The main purpose of this
software is to make genotype imputation in order to predict unobserved genotypes
from a panel with milions of haplotypes. Due to the size of the panel, the use
of \textit{PBWT} proved inevitable, further demonstrating the importance of this
data structure.\\
Durbin himself, the author of the \textit{PBWT}, with Shchur et al. in 2019,
proposed a use of his data structure in \textit{GWAS} context. Infact they
studied the associations between genetic variants in order to identify signals
of natural selection to build the so-called \textbf{ancestral recombination
  graph}, that contains complete informations about samples history.\\
For the sake of completeness, an example where \textit{PBWT} is not used is
\textbf{Ranbow}, proposed by Moeinzadeh et al. in 2020. The 
aim of this project was the haplotype reconstruction of polyploid genome from
short read sequencing data, studying also the multi-allelic case.
\subsubsection*{Research goals}
Therefore, for my master's thesis, I had to rethink the concept of
\textit{Matching Statistics} for \textit{PBWT} (tracking  a \textit{row} of the
panel instead of the \textit{pos}), how to compute the \textit{SLP}
for the panel, how to use \textit{thresholds}/\textit{LCE queries} and how to
get the same behaviour of the $\varphi$ function, in order to obtain the
\textbf{RLPBWT}, combining the ideas related to the \textit{RLBWT} with those of
the \textit{PBWT}, eventually, as for example for the $\varphi$ function,
developing a simple new data structure.\\  
% More formally, given a panel $X$ (where every row is 
% defined by $X_j$) and a pattern $P$, $|P|=X_{width}=n$, we define \textit{MS} as
% an array of length $n$, such that, for each position $1\leq i\leq n$,
% $X_{MS[i].row}[i-MS[i].len+1: i]=P[i-MS[i].len+1:i]$ and $P[i-MS[i].len:i]$ is
% not a suffix of $X_1[1:i],\ldots,X_{X_{height}}[1:i]$.
% Thresholds are defined
% by the index of the minimum \textit{LCP value} inside a run, considering also
% the \textit{LCP value} of the head of the next run, if it exists. We stretch the
% reverse panel (from the right to the left) to get the \textit{SLP}, in order to
% make \textit{LCE queries} possible. This has to be done because \textit{LCE
%   queries} are computed, between two rows, from the right to the left. Instead,
% regarding the $\varphi$  function and its inverse, I have implemented a new
% simple data structure to identify which prefix array values are above and below
% a given index in a column of the \textit{RLPBWT matrix}, that is the panel
% permuted via the prefix array and run-length encoded.
Obviously, there are some limitations, as in Durbin's \textit{PBWT}, such as
studying only biallelic panels and ignoring the management of missing data,
which are very frequent in real cases. \\
Especially during my master’s degree I have deepened both the most theoretical
issues related to algorithms, especially in the field of bioinformatics, that
the most modeling and inference issues, related to the field of computational
systems biology, so it is my interest to continue my studies with the PhD
in order to be able to deepen the computer science potential in the biological
context. Summarizing, some of the research goals of my \textit{Phd} are:
\begin{enumerate}[leftmargin=.2in]
  \setlength\itemsep{-0.2em}
  \item \textbf{Multiallelic data}. Thanks to the increasing growth in the
  production of genotype data the number of multiallelic sites is expected to
  grow, as the number of samples increases, even in the human
  genome, although now, for example, there is only a 2\% presence of triallelic
  sites. Moreover, not only could such sites be more than expected but they are
  usually not considered by most tools. A first step in this direction was made
  by Naseri with the already cited \textit{m-PBWT}. From the point of view of
  spatial complexity, especially about stored arrays for the \textit{FM-index},
  a run-length encoded version of the same could 
